---
title: 'RCE: Visualizing the Results, Part III: Wealth and Corruption'
tags: rce
output:
  html_document:
    keep_md: yes
    self_contained: no
    cache: yes
---

(This post explores some final correlations between corruption and wealth. For the rest of the posts in the series, see [here](http://politicalpathologies.blogspot.co.nz/search/label/rce). It will also be updated if any final late submissions come in -- I think there may be a couple of people who are still working on their classification. I've corrected some small errors in the previous version of the dataset).

(Last updated `r lubridate::here()`)

```{r dataLoading, echo=FALSE,warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(countrycode)
library(reshape2)
library(stringr)
data <- read.csv("rce.csv")
data_tv <- read.csv("rce_tv.csv")
fh <- read.csv("../../../Data/fh.19722014.csv")
wgi_cc <- read.csv("WGI Corruption Control Dataset.csv")
gdp <- read.table("../../../Data/gdpv6.txt",header=TRUE)

fh$country <- countrycode(fh$country, origin = "country.name", destination = "country.name", warn=TRUE)
wgi_cc$country <- countrycode(wgi_cc$country, origin = "country.name", destination = "country.name", warn=TRUE)
gdp$country <- countrycode(gdp$statenum, origin="cown",destination="country.name",warn=TRUE)

wgi_cc <- melt(wgi_cc, id.vars=1:2)
wgi_cc <- wgi_cc %>% mutate(year = as.numeric(str_extract(variable,"[0-9]+")),variable = str_extract(variable,"Estimate|StdErr|NumSrc|Rank|Lower|Upper"))
wgi_cc <- dcast(wgi_cc, country + year ~ variable)

data_tv <- left_join(data_tv,wgi_cc)

data_tv <- data_tv %>% mutate(fh = 14 - (pr + cl))

data_tv <- left_join(data_tv,gdp)

fh_gdp <- left_join(fh, gdp %>% filter(!is.na(country)))

wgi_gdp <- left_join(wgi_cc, gdp %>% filter(!is.na(country)))

wgi_gdp <- left_join(wgi_gdp, data_tv)

wgi_gdp <- wgi_gdp %>% ungroup() %>% mutate(regime_type = cut(Score,3,labels=c("Non-democratic","Hybrid","Democratic"),include.lowest=TRUE))

```

In the [last post of this series](http://politicalpathologies.blogspot.com/2015/05/rce-visualizing-results-part-ii.html), we saw that though there was a slight correlation accross countries between "being more democratic" and "being less corrupt," there was little obvious correlation between corruption and democracy over time: countries that became more democratic over the two decade period we investigated did not seem to have become obviously less corrupt, or vice-versa. We do not necessarily have a good explanation for this, but I suggested that wealth might have something to do with it. In this post, I look at correlations (across countries and over time) between corruption and wealth (as measured by GDP per capita).

The first graph shows a simple scatterplot between the *level* of income at a given point in time, per person (in constant 2005 American dollars -- that is, inflation-adjusted numbers) and the [WGI][WGI] index of corruption perceptions. Each dot represent a country-year (e.g, Singapore 2006); if a particular year was classified by POLS209, it is given a color representing the regime type the class gave it.[^Levels]

```{r, echo=FALSE,warning=FALSE,message=FALSE}
qplot(data=wgi_gdp, x= Estimate, y = rgdppc, color=regime_type,group=0) + scale_y_log10(label=dollar) + geom_smooth() + labs(y="Real GDP per capita (2005 constant $)",x="World Governance Indicators \nCorruption Control Index \n(higher is less corruption)",color = "PLS209 regime type") + theme_bw() + theme(legend.position = "bottom") + ggtitle("Wealth vs.\n corruption (World Bank Corruption Control Index)")

```

As you can see, there is a fairly strong correlation between income per capita and low corruption; almost all very rich countries (whatever their regime type) display low levels of corruption perceptions, while poorer countries tend to be seen as more corrupt. A few middle-income countries are fairly corrupt but no highly corrupt country is among the very rich (income greater than about $25,000 per capita). The richest highly corrupt country is Equatorial Guinea, a result of the sudden influx of oil money there in the 1990s. (Almost all of this money was captured by the political elite, however; most people in Equatorial Guinea are very poor).

This correlation, of course, does not tell us much about causation; we do not know if countries become rich because they happen to be less corrupt, or if they are seen as less corrupt because they are rich, or something else. A proper study of the relationship between these things would need a sophisticated statistical model in order to even begin to talk about causality, and we do not have the time or technical know-how to do that here. But we can at least take a look at the temporal relationship between corruption and income per capita in each country. The graph below is ordered from poorest to richest country, and it plots the corruption rank (grey ribbon) and the percentile rank of income (dots, colored by POLS209 regime type) on the same scale, per year:

```{r, echo=FALSE,warning=FALSE,message=FALSE,fig.height=30}
wgi_gdp <- wgi_gdp %>% filter(!is.na(rgdppc)) %>% group_by(year) %>% mutate(income_rank = percent_rank(rgdppc))
wgi_gdp$Rank <- rescale(wgi_gdp$Rank)
wgi_gdp$Lower <- rescale(wgi_gdp$Lower)
wgi_gdp$Upper <- rescale(wgi_gdp$Upper)

wgi_gdp$country <- reorder(wgi_gdp$country,wgi_gdp$income_rank)

ggplot() + geom_point(data= wgi_gdp, aes(x=year,y=income_rank,color=regime_type)) + geom_path(data= wgi_gdp, aes(x=year,y=income_rank,color=regime_type,group=0)) + geom_path(data= wgi_gdp, aes(x=year,y=Rank))+ geom_ribbon(data= wgi_gdp, aes(x=year,y=Rank, ymin=Lower, ymax=Upper),alpha=0.2) + theme_bw() + guides(color=guide_legend(title.position="top",direction="horizontal")) + labs(x="Year",y="Income rank \nand WGI corruption rank \n(rescaled to 0-1)",color="POLS209 regime type") + theme(legend.position = "bottom") + ggtitle("Income rank and corruption rank, per year") + facet_wrap(~country,ncol=4) 

```

As we can see, with some exceptions, the income rank of a country tracks quite well its corruption rank. The exceptions are nevertheless interesting. For example, Turkmenistan and Equatorial Guinea are far more corrupt than their income levels would indicate, while Rwanda has become less corrupt than its income levels would indicate. We can extract the countries with the greatest differences between income rank and corruption rank in this graph:  

All code for this post is available [in this repository][Code]. If you are technically minded and want to learn some more, you might wish to download the [R programming language](http://www.r-project.org/) and [RStudio](http://www.rstudio.com/) and play with it; I can help you if you are interested. (If you want to replicate the graphs and correlations above, you will also need a spreadsheet with the Freedom House scores, available [here](https://drive.google.com/file/d/0B5wyt4eDq98GZVhUamRKVXJpLVU/view?usp=sharing), the [WGI Corruption Control Index][WGI], available in the repository, and the [expanded dataset of GDP per capita values prepared by Gleditsch and available here][Gleditsch].)

[^Levels]: Note that the *level* of per-capita income at a given point in time is not the same thing as the rate of *growth* of income at a given point in time; poor countries (e.g., China even today -- though China is now more of amiddle-income country) can be growing quickly, and rich countries can be growing slowly (e.g., Spain today). And of course the total income of a country, divided by its population (which is more or less what GDP per capita means) doesn't say anything about how that income is distributed -- in some cases, much of that income may be captured by a small elite. But we work with the numbers we have, not the numbers we would like to have. 

[Rydland]: http://www.nsd.uib.no/rapport/nsd_rapport124.pdf
[WGI]: http://info.worldbank.org/governance/wgi/index.aspx#doc-methodology
[TI]: http://www.transparency.org/
[Code]: https://github.com/xmarquez/RCE-2015
[Polity]: http://www.systemicpeace.org/inscr/
[FH]: http://freedomhouse.org
[Gleditsch]: http://privatewww.essex.ac.uk/~ksg/exptradegdp.html